First Reproducible Research Project -- Brian Davies
===================================================

Loading and Processing the Data
-------------------------------

I'm going to cache this as it's a bit time consuming.

```{r Download and Process Data, cache = TRUE}
download.file( 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip',
               destfile = 'activity.zip', method = 'curl' )
unzip( 'activity.zip' )
activityFrameScratch <- read.csv( 'activity.csv', header = TRUE )
activityFrame <- data.frame( activityFrameScratch$date,
                             activityFrameScratch$interval,
                             activityFrameScratch$steps )
rm( activityFrameScratch )
colnames( activityFrame ) <- c( 'date', 'interval', 'steps' )
head( activityFrame )
```

Analysis of Steps Taken Per Day
--------------------------------

Get summary stats for steps per day. That function is probably a bit
more complicated than it needs to be, but never mind...

```{r Aggregate daily steps}
dailySteps <- aggregate( activityFrame$steps, list(activityFrame$date),
                         function( x ){ sum( x, na.rm =TRUE ) } )
head ( dailySteps )
```

... and plot them and summarize them ...

```{r Analyze Daily Steps}
hist( dailySteps$x, breaks = 16, col =' gray80',
      xlab = "Steps Per Day",
      main = "Daily steps frequency analysis" )
summary( dailySteps$x )
```

The output from __summary()__ is good enough for EDA purposes.

Average Daily Activity Pattern
------------------------------

Should be easy enough to calculate daily average summary using same general
strategy as daily sums ...

```{r Aggregate interval steps}
intervalStepsAvg <- aggregate( activityFrame$steps,
                            list(as.factor(activityFrame$interval)),
                            'mean', na.rm =TRUE )
length(intervalStepsAvg$Group.1)
head ( intervalStepsAvg )
```

There are `r 60/5*24` five-minute periods in a day, and the result frame
has `r length( intervalStepsAvg$Group.1 )` rows.

Analysing the results ...

```{r Analyze interval steps}
plot( intervalStepsAvg$x, type='l', xlab = 'Interval', ylab = 'Steps',
      main = 'Average steps by 5-minute interval' )
which.max(intervalStepsAvg$x)
```
Step 104 is `r (104-1)*5/60` hours into the day. The graph looks  kinda credible
i.e. almost no steps in the early morning.

Imputing Missing Values
-----------------------

Using the __is.na()__ function, there are `r sum(is.na(activityFrame$steps))`
missing step counts.

RIGHT, I just noticed that those nice little interval designators in the raw
data do things like this:

```{r funny times}
head( activityFrame$interval, 15 )
```

See how they go from 55 to 100? That means they're basically __hhmm__ with
the __hh__ omitted if possible. I can't be bothered to figure out
the time libraries so I'm going to write a little function to turn them
into minute counts to generate an index into the averages frame.

```{r define and test Interval2Minutes}
Interval2Minutes <- function( interval ){
  interval <- as.integer( interval )
  hours_part <- floor( interval / 100 )
  minutes_part <- interval - 100 * hours_part
  minutes <- hours_part * 60 + minutes_part
minutes
}
Interval2Minutes( 245 )
Interval2Minutes( c( 100, 250 ) )
Interval2Minutes( c( '100', '250' ) )
```

Just check that __intervalStepsAvg__ is in the right order. This should be zero:

```{r Check interval averages in right order}
which(as.numeric(as.character(intervalStepsAvg$Group.1))[2:288] <=
as.numeric(as.character(intervalStepsAvg$Group.1))[1:287])
```

Right, we're ready to go!

```{r Impute missing steps as average}
indexMissingSteps <- which(is.na(activityFrame$steps))
intervalMissingSteps <- activityFrame$interval[indexMissingSteps]
indexWantedSteps <- Interval2Minutes(intervalMissingSteps) / 5 + 1
wantedSteps <- as.integer(round(intervalStepsAvg$x[indexWantedSteps]))
activityFrameImputed <- activityFrame
activityFrameImputed$steps[indexMissingSteps] <- wantedSteps
```

And now we analyze the results.

```{r Aggregate & Analyze Imputed daily steps}
dailyStepsImputed <- aggregate( activityFrameImputed$steps,
                                list(activityFrameImputed$date),
                                function( x ){ sum( x, na.rm =TRUE ) } )
hist( dailyStepsImputed$x, breaks = 16, col =' gray80',
      xlab = "Steps Per Day",
      main = "Imputed daily steps frequency analysis" )
summary( dailyStepsImputed$x )
```

Weekday/Weekend Activity Patterns
---------------------------------

I built the following up after a lot of trial and error! Ternary
logical __ifelse()__ is a BIG help here.

```{r Add weekday/weekend marker}
activityFrameImputed$dayType <-
  as.factor(
    ifelse( grepl( '^S.+day$',
                   weekdays( as.Date( activityFrameImputed$date ) ) ),
            'weekend', 'weekday' )
    )
```

Now use a slightly different __aggregate()__

I've also added an __hours__ column to give the (fractional) hour of day so I
can use an __xyplot__ without the obvious gaps in the exemplar in
the briefing notes

```{r Get averages by interval for wekday and weekend, add hours column}
intervalStepsAvgByType <- aggregate( activityFrameImputed$steps,
                                     list(as.factor(activityFrame$interval),
                                          activityFrameImputed$dayType),
                                     'mean' )
intervalsHours <- seq(0,23.999,5/60)
intervalStepsAvgByType$hours <- c(intervalsHours,intervalsHours)
colnames(intervalStepsAvgByType) <- c( 'interval', 'dayType', 'steps', 'hours' )
head(intervalStepsAvgByType)
tail(intervalStepsAvgByType)
```

OK, ready to do the graph!

```{r Weekday/Weekend average interval steps plots}
library( 'lattice' )
xyplot(intervalStepsAvgByType$steps ~ intervalStepsAvgByType$hours
        | intervalStepsAvgByType$dayType,
        type='l', lwd=2, layout=c(1,2),
        xlab = 'Hour of Day', ylab = 'Average steps per 5 minutes'
        )
```

Some pretty obvious differences there, I'd say.

* Gets up later at weekends
* Lower morning peak at weekends (no walk to work?)
* More late-night activity at weekends
